### 2016-01-26 doc/thesis/intro.tex
Writing yay

A \defn{species tree} is a particular type of phylogeny in which the taxa are
species, and the branching times correspond to historical speciation events. In
general, we do not have access to the true tree relating a particular group of
extant species, as this would imply certain knowledge of speciation events in
the distant past. Therefore, all species trees considered by researchers are
estimates, based on the genetic or phenotypic similarity of the extant taxa,
the fossil record, or both. We discuss some of the challenges associated with
this estimation in subsection \ref{subsubsec:treeconv} below. However, for the
moment, we ignore these complications and consider what can be said about the
true species tree.

Speciation, which causes branching in a species tree, often occurs by an
\defn{allopatric} process, where a sub-population of organisms is isolated from
the original population by a geographic barrier~\autocite{coyne2004speciation}.
Over time, differing selection pressures on either side of the barrier, coupled
with genetic drift, cause the two populations to diverge genetically. This
eventually results in two distinct species, which will appear as two subtrees
or \defn{clades} in their species tree if descendants of both survive to the
present. Of course, this is not necessarily the case, as selection may cause
one or both of the new sub-populations or their descendants to become extinct.
On the other hand, selection may favour one of the sub-populations, allowing it
to expand into new regions or niches and possibly paving the way for subsequent
speciation events. Therefore, \comment{Is this a fair statement?} we can point
out two forces primarily responsible for the shapes of species trees:
geographic barriers, which cause branching, and selection, which controls which
lineages survive to the present and influences the branching rate.

A different type of tree used in the study of pathogens is a \defn{transmission
tree}. In such trees, tips represent infected hosts, while internal nodes
correspond to transmissions from one host to another. Transmission trees are
generally time-scaled \comment{Is this true?} and rooted, with the root
corresponding to the index case and branching times indicating times of
transmission. The internal nodes may be labeled with the donor of the
transmission pair, if this is known. The tips of the tree, rather than being
fixed at the present day, are placed at the time at which the individual was
discovered or their pathogen population sampled. Consequently, the transmission
tree may not be ultrametric, but may have tips located at varying distances
from the root. Such trees are said to have \defn{heterochronous} taxa or
samples~\autocite{drummond2003measurably}, in contrast to the
\defn{isochronous} samples found in macro-organisms species trees. A
transmission tree is illustrated in Figure~\ref{fig:contactnet} (right).

@@

The fact that these evolutionary (ie. speciation) and epidemiological processes
occur on the same time scale for RNA viruses~\autocite{drummond2003measurably}
is the basis for the field of
\defn{phylodynamics}~\autocite{grenfell2004unifying}. The similarity between
transmission and speciation means that it is possible to apply many of the
tools and techniques developed for species trees to viral transmission trees,
and interpret the results in the context of transmission rates and host
epidemiology. As a basic example, the lineages-through-time
plot~\autocite{nee1992tempo}, which plots the historical speciation rate
against time, can be used to quantify the incidence of new infections over the
course of an epidemic. \comment{Reference?} In
subsection~\ref{subsubsec:treeshape} below, we review some of these methods in
more detail, as well as several which have been developed specifically for
viral phylodynamics, and discuss what these methods can tell us about host
epidemiology.

@@

@@

@@

\subsubsection{Generative models}
\label{subsubsec:generative}

@@

>>>>

A \defn{species tree} is a particular type of phylogeny in which the taxa are
species, and the branching times correspond to historical speciation events. By
speciation events, we mean times when two formerly interbreeding groups of
organisms ceased to reproduce with each other. In general, we do not have
access to the true tree relating a particular group of extant species, as this
would imply certain knowledge of speciation events in the distant past.
Therefore, all species trees considered by researchers are estimates, based on
the genetic or phenotypic similarity of the extant taxa, the fossil record, or
both. We discuss some of the challenges associated with this estimation in
subsection \ref{subsubsec:treeconv} below. However, for the moment, we ignore
these complications and consider what can be said about the true species tree.

Speciation, which causes branching in a species tree, often occurs by an
\defn{allopatric} process, where a sub-population of organisms is isolated from
the original population by a geographic barrier~\autocite{coyne2004speciation}.
Over time, differing selection pressures on either side of the barrier, coupled
with genetic drift, cause the two populations to diverge genetically. This
eventually results in two distinct species, which will appear as two subtrees
or \defn{clades} in their species tree if descendants of both survive to the
present. Of course, this is not necessarily the case, as selection may cause
one or both of the new sub-populations or their descendants to become extinct.
On the other hand, selection may favour one of the sub-populations, allowing it
to expand into new regions or niches and possibly paving the way for subsequent
speciation events. It must be noted that speciation is not necessarily
allopatric. \defn{Sympatric} speciation refers to the emergence of a new
species without any geographic barrier~\autocite{coyne2004speciation}. Indeed,
there is a whole continuum of possible speciation
processes~\autocite{fitzpatrick2008if}, ranging from completely allopatric to
completely \defn{sympatric}. However, at least for eukaryotic organisms,
allopatric speciation is thought to be by far the most common
type~\autocite{coyne2004speciation}.

A different type of tree used in the study of pathogens is a \defn{transmission
tree}. In such trees, tips represent infected hosts, while internal nodes
correspond to transmissions from one host to another. Transmission trees
generally have branch lengths in units of calendar time, with a root
corresponding to the index case, and branching times indicating times of
transmission. The internal nodes may be labeled with the donor of the
transmission pair, if this is known. The tips of the tree, rather than being
fixed at the present day, are placed at the time at which the individual was
discovered or their pathogen population sampled. Consequently, the transmission
tree may not be ultrametric, but may have tips located at varying distances
from the root. Such trees are said to have \defn{heterochronous} taxa or
samples~\autocite{drummond2003measurably}, in contrast to the
\defn{isochronous} samples found in macro-organisms species trees. A
transmission tree is illustrated in Figure~\ref{fig:contactnet} (right).

@@

The fact that these evolutionary (ie. speciation) and epidemiological processes
occur on the same time scale for RNA viruses~\autocite{drummond2003measurably}
is the basis for the field of
\defn{phylodynamics}~\autocite{grenfell2004unifying}. The similarity between
transmission and speciation means that it is possible to apply many of the
tools and techniques developed for species trees to viral transmission trees,
and interpret the results in the context of transmission rates and host
epidemiology. As a basic example, the lineages-through-time
plot~\autocite{nee1992tempo}, which plots the historical speciation rate
against time, can be used to quantify the incidence of new infections over the
course of an epidemic~\autocite{holmes1995revealing}. In
subsection~\ref{subsubsec:treeshape} below, we review some of these methods in
more detail, as well as several which have been developed specifically for
viral phylodynamics, and discuss what these methods can tell us about host
epidemiology.

@@

A \defn{gene tree} is another type of phylogeny which traces the evolutionary
history of one or more sections of genetic material. Gene trees can be built
from small parts of genes up to whole genomes, but for ease of exposition we
will use only the word ``gene''. The tips of a gene tree represent the
different \defn{alleles}, or versions, of the gene found in the studied taxa.
The branching points in a gene tree represent genetic divergence events - that
is, the times when two distinct alleles first came into existence through
mutation.

Gene trees are usually estimated using the nucleotide or amino acid sequences
of the genes in question. Although there are methods for estimating gene trees
using other types of molecular data, such as gene orders, we will focus on
sequence-based methods here. As with species trees, it is not possible to know
gene trees with certainty, only to estimate them. However, the problem of
estimating gene trees directly from data is somewhat better defined than that
of estimating species trees, because of the quantitative nature of the
molecular data used to label the tips. In general, methods to estimate gene
trees fall into three major categories: distance-based, maximum parsimony, and
model-based~\autocite{nei2000molecular}. The first two categories are generally
no longer used for phylogenetic inference, since model-based methods have been
demonstrated to be almost universally more accurate. Therefore, we will review
in detail only the last category.

Model-based methods include both maximum likelihood and Bayesian approaches. To
use them, one must first define a parametric model of sequence evolution, here
denoted $M$. For all possible trees $T$ (with branch lengths) and choices of
model parameters $\theta$, the model must define a probability density function
on observed sequence data $D$, denoted $f(D \mid T,\,\theta)$. This is also
sometimes written $\Pr(D \mid T,\,\theta)$, even though it is a probability
\emph{density}, not a probability. Maximum likelihood aims to find a particular
tree and values for the model parameters which optimize the \defn{likelihood
function} $\L$ for the observed data,
\[
  \L(\theta,\, T \mid D) = f(D \mid T,\, \theta).
\]
Bayesian methods, rather than trying to find point estimates for $\theta$ and
$T$, aim to approximate the entire \defn{posterior distribution} for the
observed data,
\[
  f(T,\,\theta \mid D) = \frac{f(D \mid T,\,\theta)f(T,\,\theta)}{f(D)}.
\]
Here, $f(T, \theta)$ denotes the prior distribution on $T$ and $\theta$, and
$f(D)$ is the marginal density of the data (that is, the integral of the
numerator over all values of $T$ and $\theta$).

A \defn{viral phylogeny} is a particular kind of gene tree which relates the
genetic sequences of viruses carried within one or more hosts. In this work, we
consider viral phylogenies constructed from one sequence per infected
individual, so that the tips of the are associated with both the viral sequence
and the corresponding host.

@@

Many tree summary statistics have been developed to assign numerical values to
phylogenies based on their properties. One of the most widely used is Sackin's
index, which measures the imbalance or asymmetry in the tree.

@@

\subsubsection{Models for contact networks}
\label{subsubsec:generative}

Throughout this section, the variable $n$ will be used to refer to the number
of nodes in the network.

An \defn{Erd\H{o}s-R\'enyi} (ER) network~\autocite{erdos1960evolution}, also
known as a Bernoulli network, was the first network model described. It has a
single parameter, $p$, which is the probability that any particular edge is
present in the network. To construct an ER network, one simply chooses
$p \times \binom{n}{2}$ distinct pairs of nodes and draws an edge between each
pair. The average degree of a node in an ER network is equal to $p \times n$.

A \defn{Barab\'asi-Albert} (BA) network~\autocite{barabasi1999emergence},
incorproates the phenomenon of \defn{preferential attachment} seen in real life
networks. This means that new connections are more likely to involve nodes
which already have a high number of connections - popular nodes tend to become
more popular. In the formulation we consider here, BA networks have two
parameters, $m$ and $\alpha$, which are used to construct the network via the
following algorithm. The network begins as a single node, then the remaining
$n-1$ nodes are added one at a time. Each time a new node is added, it is
connected to $m$ other nodes. The probability of one of these connections
involving an existing node of degree $k$ is proportional to $k^\alpha$. Note
that in the original formulation~\autocite{barabasi1999emergence}, there was an
additional parameter $m_0$ which indicated the number of initial vertices in
the network, and the only value of $\alpha$ considered was 1. For our purposes,
we fix $m_0 = 1$ but allow $\alpha$ to vary. We refer to $\alpha$ throughout
this work as the \defn{preferential attachment power} or just attachment power.

Networks produced by the BA model are \defn{scale free}, which means that the
probability of a node having $k$ connections is proportional to $k^\gamma$ for
some constant $\gamma$. This is also called a \defn{power law} degree
distribution. An important remark on notation must be made here: in network
literature, the Greek letter $\alpha$ has often been used to refer to the power
law exponent of the degree distribution, not the preferential attachment power.
However, the paper defining the network model~\autocite{barabasi1999emergence}
uses $\gamma$ for the power law exponent, and $\alpha$ for the attachment
power, and we shall do likewise. This must be noted because there are reports
in the literature of the power law exponent being as large as 4, which would
seem to conflict with our choice (see subsection XX) to bound $\alpha$ above by
2. The reader should rest assured that it is the attachment power we are
bounding, not the power law exponent.

WS and BA networks differ from ER networks in an important way: it is possible
to explicitly calculate the likelihood of the ER model parameter $p$ given an
observed network, but this cannot be done for the other two models. This is
because the probability of a particular edge occuring is always equal to $p$ in
an ER network, regardless of the presence or absence of any other edges. In
other words, the edges can be viewed as a series of $\binom{n}{2}$ Bernoulli
trials with success probability $p$, so that the likelihood of $p$ given an
observed graph with $k$ edges is distributed as 
Binomial$\left(\binom{n}{2}, p\right)$.

@@

\subsubsection{Overview and motivation}

Mathematical models can be grouped into three categories in terms of our
ability to fit them to observed data. First, there is a limited set of models
which can be fit explicitly by using calculus to minimize a loss function. The
canonical example of this type of model is a linear regression, with the
ordinary least-squares loss function. A linear regression is a model of the
relationship between some data points $x_1, \ldots, x_n$ and outcomes $y_1,
\ldots, y_n$ which assumes that the outcomes are linearly related to the data,
modulo some noise caused by measurement errors, environmental fluctuations, and
so on. In other words, there is a constant $b$ such that 
\begin{align}
  \label{eq:regression}
  y_i = bx_i + \varepsilon_i, 
\end{align}
where $\varepsilon_i$ is the error associated with the outcome $y_i$. Notice
that, if we fix $b$, then since $y_i$ and $x_i$ are both given, $\varepsilon_i$
is also fixed. The process of fitting a linear model means choosing the
``best'' $b$. For least squares, this optimal $b$ is defined as the one which
minimizes the sum of the squared $\varepsilon_i$.

Now that we have defined what it means to fit a least-squares linear
regression, there remains the question of how to go about finding this optimal
$b$ value. Recall that the errors $\varepsilon_i$ are fixed for a particular
$b$, and our objective is to minimize the loss function, which is the sum of
the squared errors. Naively, we could simply try many different $b$ values,
calculate the loss function for each, and choose the $b$ which gives the
smallest value. This approach, though basic, falls under the umbrella of
\defn{numerical optimization}, which is the most common way the majority of
models are fit. Of course, there is a whole field devoted to more sophisticated
methods for exploring the parameter space, but they all boil down to the basic
method of trying different values and choosing the one which minimizes the
loss.

In the case of least-squares linear regression, there is another method we
could use. Looking at equation~\ref{eq:regression} for the model, it is easy to
rearrange terms to solve for $\varepsilon_i$ in terms of the other variables:
we get $\varepsilon_i = y_i - bx_i$. In fact, we can write down the entire loss
function, which is the sum of the squared $\varepsilon_i$ in terms of $b$:
\[
  \sum \varepsilon_i^2 = \sum (y_i - bx_i)^2,
\]
where the sums are over all $i$ from 1 to $n$, the number of points in the data
set. We will not describe the procedure here, but it is possible to explicitly
minimize the right-hand side by finding the point where the derivative with
respect to $b$ is zero. However, the method of directly minimizing the loss
function with calculus is only applicable to a very narrow class of simple
models. For the majority, the loss function is too complicated to set the
derivative with respect to the parameters to zero and solve.

Because of the limited space of models which permit exact solutions, numerical
optimization of a loss function is the most frequently used method to fit a
model.

An approximate Bayesian computation approach to least-squares regression might
be as follows. We will need to make an additional assumption: that the errors
$\varepsilon_i$ are normally distributed. That is, our linear model is now of
the form
\[
  y_i = bx_i + \N(0, \sigma),
\]
where $\N(0, \sigma)$ indicates a normal distribution with mean zero and
variance $\sigma$. We do not know $\sigma$ in advance, so this is another
parameter we will need to estimate. There is nothing special about normal
distributions - we could have chosen any other distribution we deemed
appropriate.

The main idea of approximate Bayesian computation is that the ``best''
parameters will define a model which produces simulated data similar to the
real data. Suppose we fix values of $b$ and $\sigma$. Then, using the given
data points $x_i$, we can generate simulated observations $y_i^*$ by adding
$bx_i$ and a randomly chosen error drawn from $\N(0, \sigma)$.

\subsubsection{Algorithms for ABC}
\label{subsubsec:abcalg}

Approximate Bayesian computation does not refer to a particular procedure for
model fitting. Rather, ABC refers to the general strategy of choosing model
parameters based on the resulting model's propensity to generate data
resembling the real data. There are three main classes of ABC algorithm which
have been developed so far: rejection, \gls{MCMC}, and sequential Monte Carlo
(SMC).

All of these approaches require some common elements. First, as with all
Bayesian methods, we are required to specify a \defn{prior distribution},
denoted $\pi$, on the parameter space. The prior specifies what we already know
or believe about the model parameters. Second, in order to compare simulated to
observed data, we need to be able to summarize a data set in a numerical
format. This is accomplished by a function, denoted $\eta$, which computes a
vector of hopefully informative summary statistics on a data set. Third, we
need a distance function $\rho$ which tells us how similar two data sets are to
each other, based on their summary statistics.

Continuing with the linear model example, we need to specify a prior $\pi(a, b,
\sigma)$ on the three parameters. We do not have much certain information about
these parameters except that $\sigma$ has to be at least zero, but it seems
reasonable to assume that extreme relationships are fairly rare, and that
positive and negative correlations are equiprobable. Therefore, we will let
$a$, $b$, and $\sigma$ be independent, $a$ and $b$ be normally distributed,
and $\sigma$ be log-normally distributed. That is,
\[
  \pi(a, b, \sigma) = 
  \begin{cases}
    \Pr[\N(0, 1) = a] \cdot \Pr[\N(0, 1) = b] \cdot \Pr[\N(0, 1) = \log\sigma]
     & \sigma > 0 \\
    0 & \sigma \le 0.
  \end{cases}
\]
For the vector of summary statistics, we will use the mean and variance of the
simulated data,
\[
  \eta(\vec{y}) = \left<\E[\vec{y}], \Var[\vec{y}]\right>.
\]
Finally, for the distance function $\rho$ we take the standard Euclidian
distance.

Rejection ABC is the simplest method, and also the one which was first
proposed~\autocite{rubin1984bayesianly}. Effectively, it comes down to the
approach described in the previous subsection of guessing parameter values
until one is close enough to the truth. More specificially, a possible set of
parameters $\theta$ is drawn from the prior distribution, and a simulated data
set $z$ is generated from the model with those parameters. If the distance
between the simulated data set and the real data, $\rho(\eta(y), \eta(z))$, is
small enough, then we accept $\theta$ as a sample from the posterior. This can
be repeated until as many samples as desired are obtained.

% MCMC

The most recently developed class of algorithms for ABC is sequential
Monte-Carlo (SMC)~\autocite{sisson2007sequential}. As with the other two
classes, we want to eventually obtain a sample from the posterior distribution
$f(\theta \mid D)$. The idea of SMC is to begin with a sample from a
distribution we know, most often the prior, and approach the posterior smoothly
by progressing through a series of intermediate distributions.

### 2016-01-26 doc/thesis/methods.tex
Writing yay

I implemented the adaptive sequential Monte-Carlo algorithm
developed in~\autocite{del2012adaptive} for approximate Bayesian computation.
In contrast to most existing sequential Monte-Carlo methods, this algorithm
does not require the user to specify a sequence of decreasing tolerances to
approach the target posterior distribution. Rather, the tolerances are computed
adaptively at each step, starting from infinity at the first iteration. The
algorithm may be stopped when the tolerance reaches a user-defined final value,
or when the rate of acceptance of the MCMC kernel reaches a user-defined
threshold. Following a heuristic applied by the
authors~\autocite{del2012adaptive}, we used the latter stopping criterion,
accepting the SMC approximation to the posterior when the MCMC acceptance rate
dropped below 1.5\%.

@@

Before undertaking a complete ABC analysis, I analysed four simple contact
network models to determine whether their parameters could be separated in tree
kernel space. The four models are described in detail in
subsection~\ref{subsubsec:generative} of the introduction. Briefly, they are:
random networks, where each possible contact has a fixed probability of
occuring; preferential attachment networks, where highly connected nodes tend
to attract more contacts; small world networks, where nodes are connected to
their immediate neighbours and the occasional far-flung contact; and full
networks, where every possible connection is present. I will describe here only
the procedure and results for the preferential attachment networks. The details
of the other three types of graph can be found in the supplemental materials.

The method of testing for separability was described previously
in~\autocite{poon2015phylodynamic}, but I will reiterate it here for
completeness. As a concrete example, consider the attachment power parameter
$\alpha$ of the preferential attachment networks. This parameter describes the
strength of attraction to highly connected nodes and is bounded below by zero
indicating no extra attraction. By qualitative observation, I determined that a
power of 2.0, which produced networks with very few ``hub'' nodes with
extremely high degree, was a suitable upper bound (see
Figure~\ref{fig:pabounds}). Therefore, I chose to test the values 0.5, 1.0, and
1.5 for separability. The other parameters were fixed: the number of nodes in
the network was 5000, and the mean degree of each node in the network was four.
As discussed in subsection~\ref{subsubsec:generative}, this is the smallest
mean degree value for preferential attachment networks which produces networks
which are more than trees.

@@

For each of the values of $\alpha$, I generated 100 networks on 5000 nodes. An
epidemic was simulated over each network (see subsection X) until 1000 nodes
were infected, and 500 of those infected nodes were sampled to form a
transmission tree. This resulted in 300 total simulated transmission trees -
100 for each of the three values of $\alpha$. Next, I computed the tree
kernel~\autocite{poon2013mapping} (see subsection~\ref{subsubsec:treeshape})
for each pair of trees. These values were placed into a 300 $\times$ 300 kernel
matrix, where the value at the $(i, j)$th position was the tree kernel of the 
$i$th and $j$th trees.

The tree kernel provides a pairwise similarity score between two trees. The
higher the kernel score, the more similar the trees are to each other.
Therefore, to have separability, we need trees simulated with the same value of
$\alpha$ to have high kernel scores with each other, but low kernel scores with
trees from different $\alpha$ values. We can visually check whether or not this
is true by laying out the trees as points on a graph, in such a way that trees
with high scores are close to each other, but trees with low scores are far
apart. This is accomplished by performing a kernel principal components
analysis (kPCA)~\autocite{scholkopf1998nonlinear} on the kernel matrix.
Briefly, ordinary principal components analysis (PCA) finds a lower dimensional
representation of points in a high dimensional space which preserves as much of
the variation in the data as possible. kPCA performs the same task, but using
the dot products of each pair of points as input, instead of the data points
themselves. A two-dimensional kPCA projection of the simulated trees is shown
in Figure~\ref{fig:pakpca}. The scenario just described - 500 samples from 1000
infected nodes - is the central panel. To ensure that the method could be used
in a variety of contexts, the same analysis was performed with 500 and 2000
infected nodes, as well as 100 and 1000 sampled tips. The projections show that
all three values are well separated from each other in many of the scenarios.
With smaller trees, it becomes harder to disinguish $\alpha = 0.5$ from $\alpha
= 1.0$ in two dimensions.

\begin{figure}[ht]
  \centering
  \label{fig:pakpca}
  \includegraphics{kernel_pa_kpca}
  \caption[Projection of kernel matrix for different attachment power values
  onto its first two principal components]{
    Projection of the kernel matrix for different preferential attachment power
    values onto its first two principal components, for eight simulation
    scenarios. Each point corresponds to a simulated transmission tree, and is
    coloured by preferential attachment power. Facets are number of infected
    nodes (horizontal), and number of sampled tips (vertical). The parameters
    to the tree kernel were $\lambda = 0.3$ and $\sigma = 5$, and the nLTT was
    not used. Qualitatively, trees with a larger number of tips are easier to
    separate in kernel space, regardless of what sampling proportion they
    represent. In all cases, the highest attachment power can be separated from
    the other two, but the two lowest values become hand to distinguish with in
    the 100-tip trees.
  }
\end{figure}

\begin{figure}[ht]
  \centering
  \label{fig:pacrossv}
  \includegraphics{kernel_pa_crossv}
  \caption[Cross-validation performance of kernel support vector machine
  classifier for preferential attachment power]{
    Cross-validation performance of kernel support vector machine classifier
    for preferential attachment power.
  }
\end{figure}

@@

@@

>>>>

I implemented the adaptive sequential Monte-Carlo (SMC) algorithm developed by
Del Moral \etal~\autocite{del2012adaptive} for approximate Bayesian
computation.

For ease of exposition, we simplifiy the notation of Del Moral \etal by
dropping the subscripts on the variables which indicate the current iteration
of the algorithm. Instead, we will add a superscript $^*$ to indicate the value
of the variable which will be used in the next iteration (this should become
clear later on). Let $\set{\theta_i}_{i=1}^{N}$ be a set of $N$ particles,
which are particular choices of the model parameters $\theta$. For each
particle $\theta_i$, let $\set{X_{i,j}}_{j=1}^M$ be a set of $M$ simulated
datasets, generated from the model with parameters $\theta_i$. Let
$\set{W_i}_{i=1}^N$ be weights associated to each particle. Let $d$ be a
distance measure on data sets, so that $d(x, y)$ is smaller the more similar
$x$ and $y$ are to each other. Let $\varepsilon$ be the tolerance level which
indicates whether a data set is ``close'' to the observed data. That is, if
$d(x, y) < \varepsilon$, we will say that $x$ and $y$ are close, otherwise they
are distant.

% begin by sampling from the prior and setting all weights to equal

Next, we calculate the next tolerance $\varepsilon^*$. Before we explain how
this is done, we first need to define how we adjust the weights on the
particles. As explained above (see subsection~\ref{subsubsec:abcalg}), the idea
of sequential Monte-Carlo is to begin with the prior distribution $\pi$,
progress smoothly through a series of intermediate distributions $\pi_1,
\ldots, \pi_{n-1}$, and eventually arrive at the target posterior distribution
$\pi_n$. In the $k$th iteration, the distribution $\pi_k$ is approximated by
the particles and their weights.

In contrast to most existing sequential Monte-Carlo methods, this
algorithm does not require the user to specify a sequence of decreasing
tolerances to approach the target posterior distribution. Rather, the
tolerances are computed adaptively at each step, starting from infinity at the
first iteration.

The algorithm may be stopped when the tolerance reaches a
user-defined final value, or when the rate of acceptance of the
Metropolis-Hastings kernel reaches a user-defined threshold. Following a
heuristic applied by the authors~\autocite{del2012adaptive}, we used the latter
stopping criterion, accepting the SMC approximation to the posterior when the
\gls{MCMC} acceptance rate dropped below 1.5\%.

@@

Before undertaking a complete ABC analysis, I analysed four simple contact
network models to determine whether their parameters could be separated in tree
kernel space. The four models are described in detail in
subsection~\ref{subsubsec:generative} of the introduction. Briefly, they are:
random networks, where each possible contact has a fixed probability of
occuring; preferential attachment networks, where highly connected nodes tend
to attract more contacts; small world networks, where nodes are connected to
their immediate neighbours and the occasional far-flung contact; and full
networks, where every possible connection is present. I will describe here only
the procedure and results for the preferential attachment networks. The details
of the other three types of graph can be found in the supplemental materials. A
graphical schematic of the analysis undertaken here is given in
Figure~\ref{fig:kernelexpt}.

\begin{figure}[ht]
  \centering
  \label{fig:kernelexpt}
  \includegraphics{kernel_expt}
  \caption[Schematic of first set of simulation experiments]{
    Schematic of first set of simulation experiments to determine separable
    parameters in kernel space, along with optimal tree kernel meta-parameters.
  }
\end{figure}

The method of testing for separability was described previously
in~\autocite{poon2015phylodynamic}, but I will reiterate it here for
completeness. As a concrete example, consider the attachment power parameter
\gls{alpha} of the preferential attachment networks. This parameter describes the
strength of attraction to highly connected nodes and is bounded below by zero
indicating no extra attraction. By qualitative observation, I determined that a
power of 2.0, which produced networks with very few ``hub'' nodes with
extremely high degree, was a suitable upper bound (see
Figure~\ref{fig:pabounds}). Therefore, I chose to test the values 0.5, 1.0, and
1.5 for separability. The other parameters were fixed: the number of nodes in
the network was 5000, and the mean degree of each node in the network was four.
As discussed in subsection~\ref{subsubsec:generative}, this is the smallest
mean degree value for preferential attachment networks which produces networks
which are more than trees.

@@

For each of the values of $\alpha$, I generated 100 networks on 5000 nodes. An
epidemic was simulated over each network (see subsection X) until 1000 nodes
were infected, and 500 of those infected nodes were sampled to form a
transmission tree. This resulted in 300 total simulated transmission trees -
100 for each of the three values of $\alpha$. The data generation steps are
shown on the left side of Figure~\ref{fig:kernelexpt}. Next, I computed the
tree kernel~\autocite{poon2013mapping} (see
subsection~\ref{subsubsec:treeshape}) for each pair of trees. These values were
placed into a 300 $\times$ 300 kernel matrix, where the value at the $(i, j)$th
position was the tree kernel of the $i$th and $j$th trees.

The tree kernel provides a pairwise similarity score between two trees. The
higher the kernel score, the more similar the trees are to each other.
Therefore, to have separability, we need trees simulated with the same value of
$\alpha$ to have high kernel scores with each other, but low kernel scores with
trees from different $\alpha$ values. We can visually check whether or not this
is true by laying out the trees as points on a graph, in such a way that trees
with high scores are close to each other, but trees with low scores are far
apart. This is accomplished by performing a kernel principal components
analysis (kPCA)~\autocite{scholkopf1998nonlinear} on the kernel matrix.
Briefly, ordinary principal components analysis (PCA) finds a lower dimensional
representation of points in a high dimensional space which preserves as much of
the variation in the data as possible. kPCA performs the same task, but using
the dot products of each pair of points as input, instead of the data points
themselves. A two-dimensional kPCA projection of the simulated trees is shown
in Figure~\ref{fig:pakpca}. The scenario just described - 500 samples from 1000
infected nodes - is the central panel. To ensure that the method could be used
in a variety of contexts, the same analysis was performed with 500 and 2000
infected nodes, as well as 100 and 1000 sampled tips.

The next step was to quantify how well trees simulated with different $\alpha$
values could be distinguished from each other. As described
previously~\autocite{poon2015phylodynamic}, this was done by assessing the
accuracy of a support vector machine regression
(SVR)~\autocite{smola1997support}. Briefly, an SVR operates by finding a
hyperplane (in two dimensions, a line) such that the deviation of most of the
data points from the line is less than some prescribed threshold. Points
further away than this threshold are ignored in the model. I performed 1000
replicate 2-fold cross-validations of an SVR predicting preferential attachment
power on the simulated trees, using the \software{ksvm} function from
\software{kernlab}~\autocite{karatzoglou2004kernlab}. That is, the SVR was
trained on a random subset of 150 trees, and then used to predict $\alpha$ of
the remaining 150 trees. The predictions were correlated against the true
values of $\alpha$ to obtain an $R^2$, and this procedure was repeated 1000
times with different subsets of trees.

The cross-validation had the dual purpose of providing a means to selecting the
optimal meta-parameters to the tree kernel (see
subsection~\ref{subsubsec:treeshape}) - they are be those which provide the
highest average $R^2$. To this end, the cross-validation was repeated for
several values of $\lambda$ and $\sigma$ as shown in Figure~\ref{fig:kernelexpt}
Each combination was evaluated with and without multiplying the tree kernel by
the normalized lineages-through-time (nLTT) statistic, and was further repeated
for the scenarios with differing numbers of infected and sampled nodes
described above. To ensure that the tree kernel was the most appropriate
similarity measure to use for ABC, we also computed the $R^2$ of $\alpha$
against Sackin's index, a widely used tree balance statistic (see
subsection~\ref{subsubsec:treeshape}), by the same cross-validation procedure.

@@

The previous set of simulations were intended to investigate which contact
network parameters could and could not be inferred by examining transmission
trees. However, they tell us nothing about the accuracy or precision we might
expect when inferring those parameters numerically. As illustrated in
Figure~\ref{fig:accprec}, the ideal situation is one where we are accurate and
precise, and the worst situation is when we are precise but not accurate.

\begin{figure}[ht]
  \centering
  \includegraphics{acc_prec}
  \caption{Illustration of accurate vs. precise kernel score estimates}
  \label{fig:accprec}
\end{figure}

Figure~\ref{fig:gridsearch} shows a schematic of this experiment. A number of
representative values, here denoted $k$, were chosen for the parameter of
interest. In the case of preferential attachment power $\alpha$, I chose $k =
8$ testing values, namely $0, 0.25, \ldots, 2.0$. For each of these values, ten
networks on 5000 nodes were generated, an epidemic of 1000 nodes was simulated
over each, and a transmission tree of 500 tips was sampled. The resulting $10
\times k$ simulated transmission trees were referred to as the \defn{testing
trees}. In addition, I chose a further $n \gg k$ training values spanning the
range of the parameter. For $\alpha$, these were $0, 0.01, \ldots, 2.0$.
Fifteen trees were simulated in the same manner for each of these values,
referred to here as \defn{training trees}.

For each of the testing trees, I computed the tree kernel with all of the
training trees. This resulted in 15 kernel scores per training value. The
training value with the highest median kernel score was used as a point
estimate of the testing value. The kernel scores were then normalized to lie in
[0, 1], and an interval was found which contained 95\% of the area under the
curve and was of minimal width. This was used as a 95\% confidence interval for
the parameter.

\begin{figure}[ht]
  \centering
  \includegraphics{gridsearch_expt}
  \caption{Schematic of grid search simulation experiments.}
  \label{fig:gridsearch}
\end{figure}

@@

Our final set of simulation experiments was to test the ABC-SMC algorithm on
simulated data. For each replicate, we generated a network on 5000 nodes,
simulated an epidemic over that network until 1000 nodes were infected, and
sampled either 500 or all 1000 of those nodes to form a transmission tree.
Priors were specified as follows: for the total number of nodes in the network,
$\Uniform(1000, 10000)$; for the number of infected nodes, $\Uniform(500,
2000)$; for the preferential attachment power, $\Uniform(0, 2)$. The mean
degree of nodes in the network was fixed at either 4, 10, or 16. The
algorithm was run with 1000 particles, and was stopped when the \gls{MCMC}
acceptance probability dropped below 1.5\%.
